Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Affective Values

### 2020-08-14 Dissonance_Santiago_v2/Dissonance_Santiago_v2.tex
v2 BLM and AP revisions

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Subjective Wellbeing},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance
from the Perspective of Subjective Wellbeing}
    \author[Some University]{Author 1\corref{Corresponding Author}}
   \ead{a1@university.com} 
    \author[Some School]{Author 2}
   \ead{a2@school.com} 
      \address[Some University]{Department, Street, City, State, Zip}
    \address[Some School]{Department, Street, City, State, Zip}

  \begin{abstract}
  \textbf{Introduction}\\
  Subjective wellbeing is a topic that has attracted considerable
  attention in recent years due to the way it correlates with health. From
  a transportation perspective, there is a burgeoning literature on the
  way travel can impact subjective wellbeing, and how this, in turn, can
  influence behavior.\\
  \textbf{Objective}\\
  The objective of this paper is to analyze a number of affective values
  associated with subjective wellbeing and the modes of transportation
  that people commonly use. In particular, we are interested in the
  potential for dissonance with respect to the primary mode of travel.\\
  \textbf{Materials and Methods}\\
  The study is based on data collected from a sample of travellers in the
  city of Santiago, in Chile. Participants in the study were asked about
  their usual mode of travel, and then were asked to name the mode or
  modes that they associate with the affective values of freedom,
  enjoyment, happiness, poverty, luxury and status. Analysis is based on
  tests of independence and visualization via mosaic plots.\\
  \textbf{Results}\\
  The results indicate that users of public transportation experience the
  most dissonance in terms of affective values, and active travellers the
  least. For those travellers who experience dissonance, active travel is
  the mode most commonly associated with freedom, enjoyment, and
  happiness, public transportation is most commonly associated with
  poverty, and the automobil is most commonly associated with luxury and
  status.

  journal: ``Journal of Transportation \& Health''
  \end{abstract}

@@

Transportation planning for decades has focused on providing mobility
for the private automobile. This is a model of development that was
initially introduced in North America as a solution to problems created
by rapid urbanization, and that was eventually copied elsewhere
(Angotti, 1996; Brown et al., 2009). Despite the initial promise of
automotive technology, it is now evident that mobility centred on the
private automobile has given rise to a litany of maladies that are in
urgent need of correction. This includes environmental concerns (i.e.,
climate change; Chapman, 2007) as well as numerous other social
(Boschmann and Kwan, 2008; Lucas, 2019, 2012), health (Khreis et al.,
2016; Milne, 2012), and equity issues (Bocarejo and Oviedo, 2012;
Martens et al., 2012; Pereira et al., 2017).

As the real impacts of our societal dependence on the private
autombobile have become increasingly evident, the transportation agenda
has aimed to shift focus to the reduction of car use and towards the
creation of mobility polycultures that offer a broader menu of
transportation alternatives than primarily (or even just) the private
automobile (Miller, 2011). In order to successfully achieve this goal,
it is essential not only to provide the services and facilities that
support public transportation and active travel, but also to attract new
users to these modes of transportation (Ettema et al., 2011). Within
this context, it has been argued that sustainable transportation
policies require all participants in the transportation system to
challenge what Gossling and Cohen (2014) termed \emph{transportation
taboos}: deep-seated ideas concerning the contribution to emissions by
individuals, the inequality of market-based approaches, and the social
and psychological functions of transportation. With respect to the
latter, it is important to move beyond a purely utilitarian focus if we
are to understand how the affective value of the transportation
experience can be leveraged to improve alternative transportation -
particularly transit and active modes (Anable and Gatersleben, 2005;
Domarchi et al., 2008; Gatersleben and Uzzell, 2007; Steg, 2005).

A useful lens to investigate the social and psychological functions of
transportation is by means of the concept of subjective wellbeing (SWB;
see, inter alia, De Vos et al., 2013; and Chatterjee et al., 2019). SWB
is defined by the OECD as ``{[}g{]}ood mental states, including all of
the various evaluations, positive and negative, that people make of
their lives, and the affective reactions of people to their
experiences'' (OECD, 2013, p. 10). As memorably put by Steg (2005): is
the use of car a must, or is it a lust? A close alignment, or
consonance, between affective values and the mode of transportation used
to travel can result in greater subjective wellbeing and increase the
probability of choosing that mode (i.e., by satisfying the \emph{lust});
on the other hand, dissonance, that is, the lack of concordance between
the mode used and affective values would be detrimental to subjective
wellbeing, especially of those for whom the use of the mode is a
\emph{must} (De Vos, 2019; Mokhtarian and Pendyala, 2018). With these
considerations in mind, one aim of the present research is to
investigate \emph{who} experiences dissonance with respect to their
primary mode of travel and affective values, namely freedom, enjoyment,
happiness, feelings of poverty, luxury, and status. Furthermore, it is
possible, even if use of a particular mode is a \emph{must}, that
travellers may still \emph{lust} for something else - in other words,
the grass may in fact look greener from the window of the car. For this
reason, we also aim to investigate which mode or modes are most commonly
associated with affective values, with a focus on those travellers who
experience dissonance in their primary mode.

Research is based on data collected from a sample of travellers in the
city of Santiago in Chile. Survey respondents were asked about their
primary mode of travel, and also about the mode or modes that they
associate with the affective values mentioned above. The paper
contributes to the literature in three ways, as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The research reported here contributes to an emerging literature on
  the topic of transportation and affective values in the context of the
  Global South (Al-Ayyash and Abou-Zeid, 2019; Bejarano et al., 2017;
  Shao and Liang, 2019; Van et al., 2014; Zorrilla et al., 2019); to the
  best of our knowledge, the case of Chile has not yet been reported.
\item
  Although there is an extensive literature on the enjoyment of commute
  and other affective values (see for instance Paez and Whalen, 2010;
  Redmond and Mokhtarian, 2001; Whalen et al., 2013; Ye and Titheridge,
  2017), from a hedonic and even eudaimonic perspectives the analysis
  has yet to be applied more fully in terms of distributional issues --
  i.e.~which groups more commonly experience dissonance (see De Vos,
  2018).
\item
  The analysis shows the attitudes of people towards their primary mode
  and their perception towards `ideal modes' -- implying their
  preferences, even in situations when their ideal mode is not part of
  their actual choice set. More concretely, the results indicate that
  users of public transportation experience the most dissonance in terms
  of affective values, and active travellers the least. For those
  travellers who experience dissonance, active travel is the mode most
  commonly associated with freedom, enjoyment, and happiness, public
  transportation is most commonly associated with poverty, and the
  automobil is most commonly associated with luxury and status. We also
  find that there are some substantial variations in dissonance by age,
  education, income, and typical commute time. The attitudes of
  travellers towards transport modes are critical factors to be
  considered by policy-makers in case they want to promote and increase
  the use of public transport or active modes (Bornioli et al., 2019; De
  Vos et al., 2019; De Vos and Witlox, 2017; Garling et al., 2019;
  Redman et al., 2013).
\end{enumerate}

The structure of the paper is as follows. After these introductory
remarks, we follow up with a background section that reviews the
literature. Next, we discuss the case study and data used in the
research. Then, the analysis and results are presented, before
concluding with some discussion and directions for future research.

@@

A consensus has emerged in the transportation community regarding the
need to complement the traditional utilitarian perspective of
transportation by looking at mobility and transport issues from the lens
of their affective functions. The affective value of transportation in
turn is important due to its potential to improve or detract from SWB.
One of the primary ways to explore this has been the satisfaction that
travelers feel towards their every day mobility experience (e.g.,
Cecilia Jakobsson Bergstad et al., 2011). As a consequence, there is a
wealth of research on satisfaction with the use of different modes of
transportation. For example, numerous studies report that car users
often have a higher level of satisfaction compared to other transport
modes (C. J. Bergstad et al., 2011; Eriksson et al., 2013; Redmond and
Mokhtarian, 2001; Whalen et al., 2013; but see Handy and Thigpen, 2019).
In a similar way, there are multiple reports that active travel also
tends to yield high levels of satisfaction (Gatersleben and Uzzell,
2007; Handy and Thigpen, 2019; Paez and Whalen, 2010; Smith, 2017;
St-Louis et al., 2014; Whalen et al., 2013), In contrast, public
transport users tend to assess their experience more negatively (Abenoza
et al., 2017; De Vos et al., 2016; Gatersleben and Uzzell, 2007; Handy
and Thigpen, 2019; Paez and Whalen, 2010). Multi-modal trips also
influence satisfaction levels; for instance, when an individual chooses
a particular mode of transportation, she will report a higher level of
satisfaction with that chosen mode -- perhaps as a form of \emph{post
hoc} validation (Susilo and Cats, 2014).

While the use of travel satisfaction has been mainly used in the context
of daily trips -- typically linked to cost-benefit and utilitarian
measurements --, the evaluation of Subjective Wellbeing (SWB) over time
has risen as an alternative measure. In the field of travel behaviour,
Ettema et al.~(2010, p. 725) define SWB as the degree to which an
individual positively evaluates the overall quality of their lives,
where the general life satisfaction encompasses a more extended
temporality -- which implies assuming a tendency to be more stable over
time. This concept has prompted a growing literature that complements
and applies SWB in a broader range of satisfaction scales and
situations. The definition of other factors such as travel choice mode,
attitudes and external elements of the built environment have been
studied for a broader understanding of the changes produced in the SWB
(e.g., Handy and Thigpen, 2019). As these factors do not necessarily
apply to the general life satisfaction on the long term, the studies
have aimed to determine both the direct and indirect effects on the
perception of users (see, e.g., Ye and Titheridge, 2017). Other concepts
have also emerged as the Satisfaction with Travel Scale (STS), a
measurement devised by Ettema et al.~(2011), as well as different scales
based on people's travel perceptions. De Vos et al.~(2015), for
instance, explore in detail the underlying dimensions of the affective
domain of STS on which SWB is based (for more on STS see also Friman et
al., 2013).

Recent literature on SWB and its link with transport have demonstrated a
relationship between people's perceptions and satisfaction with their
daily travel (e.g., Smith, 2017; Mokhtarian and Pendyala, 2018; St-Louis
et al., 2014). Scholars have shown that accessibility has been the most
developed factor that influences people's wellbeing (Delbosc, 2012), and
activities have a direct impact on travel satisfaction (Cecilia
Jakobsson Bergstad et al., 2011). Delbosc (2012, p. 28), for instance,
has summarised the most significant influences on psychological
wellbeing: poverty and employment, meaningful relationships and health.
However, understanding the components affecting people's perceptions
implies the differentiation between affective (also named as
symbolic-affective) and instrumental values (C. J. Bergstad et al.,
2011). Steg et al.~(2011) have compared symbolic-affective opposed to
instrumental-reasoned motives based on car-use, and other studies have
also found associations between affective and symbolic aspects of
car-use (see, e.g., Gatersleben and Uzzell, 2007; Lois and Lopez-Saez,
2009). Previous studies have also demonstrated how socio-demographic
factors affect the levels of SWB. The effect of income on SWB (Clark and
Oswald, 1996; Ferrer-i-Carbonell, 2005); education and unemployment
(Argyle et al., 1999); age (Diener and Eunkook Suh, 1997), and gender
(Tesch-Römer et al., 2008) have been extensively studied. Recent
research also suggests the links between commuting, SWB and emotional
wellbeing assessment (\textbf{Olson et al.~2013}; \textbf{Kahneman et
al.~2004}). However, more research is needed to understand how these
socio-demographic variables connect as well with the affective responses
to mode of travel (St-Louis et al., 2014).

The research needs already recognized in the developed world are also
markedly acute in the context of the Global South, where historical
inequality has tended to create a symbolic attachment to the automobile,
in addition to negative connotations for public transport and active
travel (Zorrilla et al., 2019). In this way, there is an emerging
literature that investigates affective factors in travel behaviour in a
number of developing countries. A cross-country study in Asia revealed
that the affective factors of public transportation and car use are
important, and in particular the social orderliness of transit was
suggested as a way to make this mode more attractive to users (Van et
al., 2014). In terms of active travel, a study in China found that
attitudes that embrace new styles and technologies despite their cost
are associated with the intention to continue using shared bicycles
(Shao and Liang, 2019). The importance of affective factors for policy
and planning is further highlighted by research in Colombia that shows
how users felt proud using a bicycle shared system, in addition to
experiencing feelings of belonging to a civic culture and the enjoyment
and pleasure of travel itself (Bejarano et al., 2017). This paper
contributes to further our understanding of affective values in travel
behaviour in a developing country.

@@

The case study to contextualize the discussion is Santiago de Chile, the
capital of one of the countries with the highest levels of inequality in
terms of socio-economic distribution in the world. These inequities have
manifested in a disproportionate expense in transport by the different
socioe-conomic sectors, a disproportion in travel times and distances
travelled. On the other hand, car ownership and use of public transport
in the capital are directly conditioned by strong spatial segregation --
the higher the income, the higher is the use of the automobile;
conversely, and the lower the income, the higher is the predisposition
to use public transport. Although transport is a reflection of deep
inequities in various dimensions, it has triggered an exceptional
discomfort and disagreement with the daily travel experience and quality
of life.

\hypertarget{sample}{%
\subsection{Sample}\label{sample}}

The study is based on a survey conducted in the city of Santiago during
the months of November and December 2016, that is, the end of the Spring
and beginning of Summer. The survey collected information on a wide
range of travel and related issues. The data collection considered a
quota-sampling method for gathering the information, considering the
socio-demographic information from Pre-Census of 2012. An equal
representation of both genders and a representation of the proportion of
inhabitants per area were chosen as relevant characteristics of the
sample. In total, there were \(n=451\) valid surveys, although not every
survey was complete and there were missing responses for some answers.

The survey was structured in eight sections, as follows: 1) Individual
characteristics of respondent; 2) Health; 3) Feelings and emotions; 4)
Reasons for travel and planning travel; 5) Social interaction; 6) Nature
and sustainability; 7) Information and telecommunications; 8) Built
environment; and 9) Commuting. For the present study, we draw data from
sections 1), 3) and 9). In terms of individual characteristics of the
respondents and their commute, participants were asked about basic
socio-demographic information, including age, level of education,
income, and the typical duration of their regular commute. The
descriptive statistics of the sample appear in Figure
\ref{fig:descriptive-statistics}. The sample tends to be younger, and
well-educated, with an almost uniform distribution of income levels. The
trend in typical commute time is towards longer commutes.

@@

In addition, respondents were asked about their primary mode of travel
for their regular commute. The modes available were Car, Taxi, Colectivo
(a form of shared ride, intermediate in flexibility and capacity between
taxi and bus); Motorcycle; Metro; Bus; Bicycle; Walking. As seen in the
top panel of Figure \ref{fig:primary-mode-travel}, the three most common
modes of travel are Metro, Bus, and Car, followed by Walking and Biycle.
For the analysis, we aggregate these modes into the following categories
(bottom panel of Figure \ref{fig:primary-mode-travel}): Car, Active
(Walking + Bicycle), Public (Metro + Bus), and Other (Taxi + Colectivo +
Motorcycle).

@@

Of particular interest for the present study is the following question
in Part 3) of the survey:

@@

To investigate the first question in our analysis, we create contingency
tables that tabulate the frequency of dissonance with respect to each
affective value, stratified by the attributes of Age, Education, Income,
Primary Mode of Travel, and typical Commute Time. Table
\ref{tab:cross-tabulation-results} presents the frequency (in
percentage) of dissonance, in other words, the percentage of respondents
out of the total in their stratum who indicated mode or mode(s) for the
affective value that do not correspond to their primary mode of travel.

>>>>

  \journal{Journal of Transport \& Health} % Sets Journal name

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Affective Values},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance
from the Perspective of Affective Values}
    \author[Some University]{Author 1\corref{Corresponding Author}}
   \ead{a1@university.com} 
    \author[Some School]{Author 2}
   \ead{a2@school.com} 
      \address[Some University]{Department, Street, City, State, Zip}
    \address[Some School]{Department, Street, City, State, Zip}

  \begin{abstract}
  \emph{Introduction}\\
  Subjective wellbeing is a topic that has attracted considerable
  attention in the transportation literature in recent years. As a result,
  there is a burgeoning literature that investigates the impacts of travel
  on subjective wellbeing, and how wellbeing, in turn, can influence
  behavior. An important aspect of subjective wellbeing are the affective
  reactions of people to their experiences.\\
  \emph{Objective}\\
  The objective of this paper is to analyze the affective reactions of
  travelers with respect to various modes of transportation. In
  particular, we are interested in the potential for dissonance between
  primary mode of travel and the mode(s) of travel identified as evoking
  various affective reactions.\\
  \emph{Materials and Methods}\\
  The study is based on data collected from a sample of travelers in the
  city of Santiago, in Chile. Participants in the study were asked about
  their usual mode of travel, and then were asked to name their ideal
  mode(s) of transportation from the perspective of various affective
  reactions. The reactions we investigate are associated with the values
  of freedom, enjoyment, happiness, poverty, luxury, and status. Analysis
  is based on tests of independence and visualization techniques.\\
  \emph{Results}\\
  The results indicate that users of public transportation experience the
  most dissonance in terms of affective reactions, and active travelers
  the least. For those travelers who experience dissonance, active travel
  is the mode most commonly associated with freedom, enjoyment, and
  happiness, whereas public transportation is most commonly associated
  with poverty. The automobile, in contrast, is the mode most commonly
  associated with luxury and status.
  \end{abstract}

@@

Transportation planning for decades has focused on providing mobility
for the private automobile. This is a model of development that was
initially introduced in North America as a solution to problems created
by rapid urbanization, and that was eventually copied elsewhere
(Angotti, 1996; Brown et al., 2009). Despite the initial promise of
automotive technology, it is now evident that mobility centered on the
private automobile has given rise to a litany of maladies that are in
urgent need of correction. This includes environmental concerns (i.e.,
climate change; Chapman, 2007) as well as numerous other social
(Boschmann and Kwan, 2008; Lucas, 2019, 2012), health (Khreis et al.,
2016; Milne, 2012), and equity issues (Bocarejo and Oviedo, 2012;
Martens et al., 2012; Pereira et al., 2017).

As the impacts of our societal dependence on the private automobile have
become increasingly evident, the transportation agenda has aimed to
shift focus to the reduction of car use and towards the creation of
mobility polycultures that offer a broader menu of transportation
alternatives other than primarily (or even just) the private automobile
(Lavery et al., 2013; Miller, 2011). In order to successfully achieve
this goal, it is essential not only to provide the services and
facilities that support public transportation and active travel, but
also to attract new users to these modes of transportation (Ettema et
al., 2011). Within this context, it has been argued that sustainable
transportation policies require all participants in the transportation
system to challenge what Gossling and Cohen (2014) termed
\emph{transportation taboos}: deep-seated ideas concerning the
contribution to emissions by individuals, the inequality of market-based
approaches, and the social and psychological functions of
transportation. With respect to the latter, subjective wellbeing (SWB)
is increasingly recognized as a policy goal in and of itself (Diener et
al., 2015; Dolan and Metcalfe, 2012), and also as an intermediate
objective to achieve ulterior goals. Accordingly, there is a growing
consensus in the field of transportation about the need to move beyond a
purely utilitarian focus to understand the affective and wellbeing
values of the transportation experience (e.g., Anable and Gatersleben,
2005; De Vos et al., 2013; Domarchi et al., 2008; Gatersleben and
Uzzell, 2007; Steg, 2005). Consideration of factors related to
subjective wellbeing, for example, are useful for policy-makers to
enhance the experience of travel, to increase satisfaction with travel,
and to meet the preferences of travelers (Chatterjee et al., 2019).

A definition of subjective wellbeing is as ``{[}g{]}ood mental states,
including all of the various evaluations, positive and negative, that
people make of their lives, and the affective reactions of people to
their experiences'' (OECD, 2013, p. 10). As memorably put by Steg
(2005): is the use of a car a must, or is it a lust? A close alignment,
or consonance, between affective reactions and the mode of
transportation used to travel can result in greater subjective wellbeing
and increase the probability of choosing that mode (i.e., by satisfying
the \emph{lust}); on the other hand, dissonance, namely the lack of
concordance between the mode used and the affective reaction to that
mode, would be detrimental to subjective wellbeing, especially to those
for whom the use of the mode is a \emph{must} (De Vos, 2019; Mokhtarian
and Pendyala, 2018). With these considerations in mind, the present
research aims to investigate \emph{who} experiences affective dissonance
with respect to their primary mode of travel. We focus our investigation
on reactions from the perspective of feelings of freedom, enjoyment,
happiness, poverty, luxury, and status - concepts termed \emph{affective
values} hereafter. Furthermore, it is possible, even if use of a
particular mode is a \emph{must}, that travelers may still \emph{lust}
for something else - in other words, the grass may look greener from the
window of a car. For this reason, we also aim to investigate which mode
or modes are most commonly identified as ideal from the perspective of
the aforementioned affective values.

Our research is based on data collected from a sample of travelers in
the city of Santiago in Chile. The analysis sheds light on the affective
reactions of people towards their primary mode of travel, and their
perceptions towards `ideal modes'. This reveals important differences in
situations when the ideal mode of a traveler is not the mode they
actually use - or in other words, the affective dissonance between
actual and ideal modes of travel. More concretely, the results indicate
that users of public transportation experience the most dissonance,
followed by automobile users, and finally active travelers, who tend to
experience the least affective dissonance. For those travelers who
experience dissonance, active travel is the mode most commonly
associated with freedom, enjoyment, and happiness, whereas public
transportation is most commonly associated with poverty, and the
automobile is most commonly associated with luxury and status. We also
find that there are significant variations in dissonance by age,
education, income, and typical commute time. The reactions of travelers
towards various transport modes are critical factors that policy-makers
need to consider to promote and increase the use of public transport and
active modes (Bornioli et al., 2019; De Vos et al., 2019; De Vos and
Witlox, 2017; Garling et al., 2019; Redman et al., 2013).

After these introductory remarks, the paper follows up with a review of
the literature on the topics of subjective wellbeing and affective
values. Next, we discuss the case study, as well as the methodology and
data used for this research. The analysis and results of the study are
presented afterwards, before concluding with some discussion and
directions for future research and policy.

Please note that this document was prepared using R Markdown and is an
example of reproducible research (Brunsdon and Comber, 2020). The R
Markdown file, along with the data file needed to reproduce the
analysis, are available for
download\footnote{\url{https://drive.google.com/open?id=189ZvfVvRis5xZA9IlviKCLW2bwbPJAgg}}.

@@

A consensus has emerged in recent years regarding the need to complement
the traditional utilitarian perspective of transportation by
investigating mobility and transport from the lens of their affective
functions. The affective value of transportation is important due to its
potential to improve or detract from SWB.

One of the primary ways to explore issues of subjective wellbeing in
transportation has been the satisfaction that travelers feel towards
their everyday mobility experience (e.g., Cecilia Jakobsson Bergstad et
al., 2011). As a consequence, there is a wealth of research on
satisfaction with the use of different modes of transportation. For
example, numerous studies report that car users often have a higher
level of satisfaction compared to other transport modes (C. J. Bergstad
et al., 2011; Eriksson et al., 2013; Redmond and Mokhtarian, 2001),
although others report highest commute satisfaction for bicycle and
train commuters (e.g., Whalen et al., 2013; Handy and Thigpen, 2019). In
a similar way, there are multiple reports that active travel also tends
to yield higher levels of satisfaction (Gatersleben and Uzzell, 2007;
Handy and Thigpen, 2019; Paez and Whalen, 2010; Smith, 2017; St-Louis et
al., 2014; Whalen et al., 2013), In contrast, public transport users
often assess their experience more negatively (Abenoza et al., 2017; De
Vos et al., 2016; Gatersleben and Uzzell, 2007; Handy and Thigpen, 2019;
Paez and Whalen, 2010). Multi-modal trips also influence satisfaction
levels; for instance, individuals tend to report higher levels of
satisfaction with the mode they use -- perhaps as a form of \emph{post
hoc} validation of their choices or circumstance (Susilo and Cats,
2014).

While travel satisfaction has often been used in the context of daily
trips -- typically linked to cost-benefit and utilitarian measurements
--, the evaluation of subjective wellbeing (SWB) over time has risen as
an alternative measure. Ettema et al.~(2010, p. 725) define SWB as the
degree to which an individual positively evaluates the overall quality
of their lives, where the general life satisfaction encompasses a more
extended temporal period -- which implies greater temporal stability.
This concept has prompted a growing literature that complements and
applies SWB in a broader range of satisfaction scales and situations.
The definition of other factors such as travel choice mode, attitudes,
and external elements of the built environment have been studied for a
broader understanding of the changes produced in the SWB (e.g., Handy
and Thigpen, 2019). As these factors do not necessarily apply to
longer-term general life satisfaction, other studies have aimed to
determine both the direct and indirect effects on the perception of
users (see, e.g., Ye and Titheridge, 2017). Other concepts have also
emerged, including the Satisfaction with Travel Scale (Ettema et al.,
2011), as well as different scales based on people's perceptions of
travel. De Vos et al.~(2015), for instance, explore in detail the
underlying dimensions of the affective domain of STS on which SWB is
based (for more on STS also see Friman et al., 2013).

The literature on SWB has demonstrated a relationship between people's
perceptions and satisfaction with their daily travel (e.g., Smith, 2017;
Mokhtarian and Pendyala, 2018; St-Louis et al., 2014). Scholars have
shown that accessibility is among the most studied factors that
influences subjective wellbeing (Delbosc, 2012), and activities have a
direct impact on travel satisfaction (Cecilia Jakobsson Bergstad et al.,
2011). Delbosc (2012, p. 28), for instance, has summarized the most
significant influences on psychological wellbeing: poverty and
employment, meaningful relationships and health. However, understanding
the components that affect these perceptions implies the differentiation
between affective (also named as symbolic-affective) and instrumental
values (C. J. Bergstad et al., 2011). Steg et al.~(2011) have compared
symbolic-affective and instrumental-reasoned motives for car-use, and
similar findings are reported in other studies (e.g., Gatersleben and
Uzzell, 2007; Lois and Lopez-Saez, 2009). Previous research also
demonstrates how socio-demographic factors affect the levels of SWB,
includin income (Clark and Oswald, 1996; Ferrer-i-Carbonell, 2005),
education and unemployment (Argyle et al., 1999), age (Diener and
Eunkook Suh, 1997), and gender (Tesch-Römer et al., 2008). However,
further research beyond the determinants of trip satisfaction is needed
to understand how these socio-demographic variables connect with the
affective reactions to various modes of travel (St-Louis et al., 2014).

The research needs outlined above are well-recognized in the developed
world, but there is still a dearth of research in the context of the
Global South. Historical inequality in many developing countries has led
to strong symbolic attachments to the automobile, in addition to
negative connotations for public transport and active travel (Zorrilla
et al., 2019). To address this gap in knowledge a body of literature has
emerged to investigate the affective aspects of travel behavior in a
number of developing countries. A cross-country study in Asia revealed
that the affective factors of public transportation and car use are
important, and in particular the social orderliness of transit was
suggested as a way to make this mode more attractive to users (Van et
al., 2014). In terms of active travel, a study in China found that
attitudes that embrace new styles and technologies despite their cost
are associated with the intention to continue using shared bicycles
(Shao and Liang, 2019). The importance of affective factors for policy
and planning is further highlighted by research in Colombia that shows
the pride users feel when using a bicycle shared system. This affective
reaction is above and beyond other positive feelings, such as a sense of
belongingness to a civic culture, and the enjoyment and pleasure of
active travel itself (Bejarano et al., 2017).

@@

The case study for the present research is Santiago. Santiago is the
capital of Chile, a country with one of the highest levels of
socio-economic inequality in the world. A tangible manifestation of the
inequalities experienced by many in Chile is the large disparity in the
relative cost of transportation, travel time, and distance traveled by
different socio-economic segments of the population. To further
complicate matters, strong spatial segregation also conditions car
ownership and the use of public transportation -- the higher the income,
the higher the use of the automobile; conversely, the lower the income,
the higher the reliance on public transportation. Although the
transportation experience is but one of many dimensions of inequality,
the experiences in this sector have triggered exceptional discomfort and
dissatisfaction among the public. Recent social unrest, triggered by a
seemingly minor hike in the fare of public transportation, has brought
many of these concerns to the forefront of the public conscience in
dramatic fashion (e.g., Davies, 2019; {\textbf{???}}; {\textbf{???}}).

Previous research has helped to contextualize everyday mobility in
Santiago relative to other Latin American cities (e.g., Avellaneda and
Lazo, 2011; Rodríguez Vignoli, 2008), but much remains to be explored.
Measuring instruments and new methods have led to more accurate and
precise understandings of the social issues that arise as consequence of
transport infrastructure and housing provision (Cox and Hurtubia, 2016),
and the minimum provision of basic services (Tiznado-Aitken et al.,
2016). However, the focus on accessibility as a measure of inequalities
remains predominant ({\textbf{???}}; Martínez et al., 2018; Niehaus et
al., 2016; Rojas et al., 2016).

\hypertarget{data}{%
\subsection{Data}\label{data}}

The study is based on a survey conducted in the city of Santiago during
the months of November and December of 2016, at the end of the Spring
and beginning of Summer seasons. The survey collected information on a
wide range of transportation and related issues, and the data collection
protocol considered a quota-sampling method for gathering the
information, considering the socio-demographic information from
Pre-Census of 2012. The survey was carried out face-to-face in centers
of activity with dense provision of offices, services, and educational
centers. An equal representation of both genders and a representation of
the proportion of inhabitants per area were chosen as relevant
characteristics of the sample. In total, there were \(n=451\) valid
reponses, although not every response was complete and some questions
have missing values.

The study considers data from 3 out of 8 sections of a longer survey.
The relevant sections of the survey concern the individual
characteristics of respondents, their feelings and affective responses
related to their commute, and aspects describing their regular commute
trips. In terms of individual characteristics and their commute,
participants were asked socio-demographic information, including age
(coded in three classes: younger than 35, older than 35 but younger than
55, and 55 and olde); level of education (three classes: Kindergarden to
grade 12, or K-12, technical diploma or university graduate, and
graduate degree), income (three classes, by tertile), and the typical
duration of the respondent's regular commute (four classes, by duration
in minutes). The descriptive statistics of the sample appear in Figure
\ref{fig:descriptive-statistics}. The sample trends to younger (56\% is
younger than 34 years old) and well-educated respondents (68\% of
respondents having technical or professional education), with an almost
uniform distribution of income levels. The trend in typical commute time
is towards longer commutes -- e.g.~55\% of the trips are longer than 40
minutes, and from those 56\% corresponds to trips over the 60 minutes
long.

@@

In addition, respondents were asked about their primary mode of travel
for their regular commute. The modes available were Car; Taxi;
\emph{Colectivo} (a form of shared ride, intermediate in flexibility and
capacity between taxi and bus); Motorcycle; Metro; Bus; Bicycle; and
Walking. As seen in the top panel of Figure
\ref{fig:primary-mode-travel}, the three most common modes of travel are
Metro, Bus, and Car, followed by Walking and Bicycle. For the analysis,
we aggregate these modes into the following categories (bottom panel of
Figure \ref{fig:primary-mode-travel}): (26\% of total modes -- which
corresponds to 91\% of the total of private motorized modes), Active
(9\% - Walking + Bicycle), Public (60\% - Metro + Bus), and Other
motorized modes (6\% - Taxi + Colectivo + Motorcycle).

@@

Of particular interest for the present study is the following question
in Part 3 of the survey:

@@

We begin by profiling the travelers who experience dissonance. The null
hypothesis is that there are no systematic differences in terms of who
tends to experience affective dissonance with respect to their primary
mode of travel. To investigate this hypothesis we create contingency
tables to tabulate the frequency of dissonance with respect to each
affective value, stratified by the individual attributes of respondents.
Table \ref{tab:cross-tabulation-results} presents the frequency (in
percentage) of dissonance. This is the percentage of respondents out of
the total in their stratum who selected, for each affective value, a
mode or mode(s) of transportation that do not correspond to their
primary mode of travel.

As seen in the table, there are five individual attributes that we
analyze. Three of these are socio-economic and demographic, namely age,
level of education, and income. The other two are transportation
related, namely primary mode of travel and commute time. The frequency
tables are tested in every case using \(\chi^2\) tests of independence
(\(p\)-values are reported in the table; lower \(p\)-values mean that
the null hypothesis of independence can be rejected with greater
confidence). It is interesting to note that the only category for which
all affective values are significant at better than 5\% level of
confidence is mode of travel.
